---
title: "TACTIC analysis script"
author: "Vallo Varik, Alessio Yang, Nicolai Karcher"
date: "2025-02-03"
output: html_document
---

# Get ready 


```{r setup, include=F}
pacman::p_load(data.table, tidyverse, ggrepel, GGally, scales, ComplexHeatmap, limma, Biobase, cowplot, janitor, yaml)
options(datatable.print.nrows = 20)
source("./helper.R")
```

# Load configuration YAML file

This file specifies general parameters assumed in this script, such as thresholds for analysis, negative control gene name etc

```{r}
# Example usage
load_yaml_to_global('./input/config.yaml')
```

# Specify replicate information

Experimental design can be complicated, with biological replicates and nested technical replicates. 
Below, I implement sanity checks that specify the expected relationship/number of replicates in the experiment.
This will be used for sanity checks down the line

```{r}
# biological replicates
biol_replicate_column_name <- 'biorep96'
expected_num_biol_replicates <- 2

# technical replicates
tech_replicate_column_name <- 'techrep96'
#expected_num_tech_replicates <- 2
# If a vector, it means number of technical replicates can vary
expected_num_tech_replicates <- c(2, 3)

# plate replicate
plate_replicate_column_name <- "plate_replicate"
# Some systems have 2, some 3 plate replicates
# TODO: This would be better solved by a mapping of systems to plate replicates, but ok for now
expected_num_plate_replicates <- c(2, 3)

```

# Specify experimental design

This code block does 2 things
- It loads the initial 96 well plate pipeting scheme, relating genes to their positions in the plate. It then maps individual wells of the 96 well plate to wells in the 384 well plate, and then to the 1536 well plate.
- Furthermore, and importantly, the registerQuadrants function specify which plate are biol. replicates, and how these are later on further technically replicated in 384 well plates

It is important to note that the final replication step (what I call 'plate replicate') is not encoded here, but instead encoded in the iris file names (see 'tail' in the config.yaml)

```{r}
# Plate maps -------------------

# 1) create/adjust the 96w maps at './input/maps/library_plate_n.csv'

# 2) go from 96w to 384w to 1536
# TODO: Document the logic here somehow.
# first 384w plate
# second 384w plate
# third 384w plate
# fourth 384w plate
plate_number_vector <- c(
  1, 2, 3, 2, 
  4, 5, 6, 6,
  1, 2, 3, 1,
  4, 5, 6, 5
  )
bio_rep_vector <- c(
  1, 1, 1, 1,
  2, 2, 2, 2,
  1, 1, 1, 1,
  2, 2, 2, 2
)
tech_rep_vector <- c(
  1, 1, 1, 2,
  1, 1, 1, 2,
  2, 3, 2, 3,
  2, 2, 3, 3
)

rep_information <- data.frame(
  plate_number = plate_number_vector,
  bio_rep = bio_rep_vector,
  tech_rep = tech_rep_vector
) %>%
  as_tibble()

print(str_c("We array a total of ", length(plate_number_vector), " 96 well plates with the following counts"))
tmp <- table(plate_number_vector)
names(tmp) <- str_c("Plate ", names(tmp))
print(tmp)



# Some diagnostics
stopifnot(length(bio_rep_vector) == length(tech_rep_vector))
if (length(unique(table(bio_rep_vector))) != 1) {
  stop("Number of biological replicates is not consistent across 96 well plates, check how you do define bio_rep_vector")
} else {
  print(str_c("Number of biological replicates (inferred from bio_rep_vector): ", length(unique(bio_rep_vector))))
  num_biol_repl <- length(unique(bio_rep_vector))
}

print("Overall, I'm seeing this:")
tmp <- rep_information %>%
  group_by(plate_number, bio_rep) %>%
  tally() %>%
  rename(tech_reps = n) %>%
  print()
w <- FALSE
if (length(unique(tmp$tech_reps)) != 1) {
  w <- TRUE
  warning("Number of technical replicates is not consistent over plates; this is possible but please make sure this is intended.")
}
if (w) {
  # seems idiotic but try putting it in the condition :)
  Sys.sleep(2)
}

```


```{r}
map96to384quadrants = registerQuadrants('384w', 
  plate_number = plate_number_vector,
  bio_rep = bio_rep_vector,
  tech_rep = tech_rep_vector
)

map384to1536quadrants = registerQuadrants("1536w",
  # TODO: TO be consistent with google scheme, make this work with letters
  plate_number = c(1, 2, 3, 4) 
) # TODO: For safety reasons, maybe remove biorep384, techrep384???
```


```{r}
folders = list.files(iris_input_folder, full.names=T)

iris = lapply(folders, loadIrisFiles) %>%
  setNames(gsub(paste0(iris_input_folder, '/'), '', folders)) %>%
  rbindlist(idcol='folder') %>%
  group_by(folder) %>%   # to annotate the same way in each folder
  mutate(plt1536 = match(file, unique(file))) %>%
  separate(file, column_names, sep='-') %>%
  mutate(tail = gsub('.JPG.iris', '', tail)) %>%
  separate(tail, c('system_num', 'plate_replicate')) %>% # This line generates the 'plate replicate'
  rename(row1536=row, col1536=column) %>%
  left_join(fread('./input/maps/systems.csv', colClasses='character')) 

iris = left_join(map96to384quadrants, map384to1536quadrants, relationship = "many-to-many") %>%
  add384RowAndCol() %>%
  add1536RowAndCol() %>%
  right_join(iris) %>%
  as_tibble() %>%
  left_join(read96wMaps()) %>%
  mutate(
    colony_id  = interaction(plt1536, row1536, col1536),
    biorep_all = interaction(across(contains('biorep'))) %>% as.numeric() %>% str_pad(2, pad='0')) %>%
  setDT() %>%
    as_tibble()

```

# Clean up data

```{r}
iris <- iris %>%
  select(
    cond, numb, genename, system_desc, opacity, colony_id, folder,
    all_of(c(biol_replicate_column_name, tech_replicate_column_name, plate_replicate_column_name))
  ) %>%
  # TODO: Rename biol_repliate_column_name to bio_rep and tech_replicate_column_name to tech_rep
  as_tibble()
```

```{r}
# Sanity check: iris object must not contain any NAs at this step
if (any(is.na(iris))) {
  stop("Collated results (iris object) contain NAs. This probably means something could not be joined in quadrant table joining. Please check the input files and ensure map files/configuration is correct.")
}
```


# QC and alternative analysis

```{r}
limma_results <- parallel::mclapply(unique(iris$folder), function(folder) {
  
  out_folder = paste0('./output/', folder)
  if(!dir.exists(out_folder)) dir.create(out_folder, recursive = T)

  fold = iris %>% filter(folder == {{folder}})
  reps = c(
    biol_replicate_column_name,
    tech_replicate_column_name,
    plate_replicate_column_name
  )

  # split to wide  ---------------------
  dat_wide = lapply(reps, function(x) {
    form = as.formula(paste('... ~', x))
    fold[[x]] = paste0('rep', fold[[x]])
    fold %>%
      # summarize gfp controls for there are hundreds of reps
      .[, .(opacity = median(opacity)), by=c('genename', 'cond', 'numb', x)] %>%
      dcast(form, value.var = 'opacity')
  }) %>% setNames(reps) 


  # plot-n-store the rep cor plots
  lapply(reps, function(x) {
    ggsave(paste0(out_folder, '/qc_', x, '_correlation.pdf'), 
      plotReplicateCorrelation(dat_wide[[x]]), h=12, w=12
    )
  })


  # plot-n-store the hierarchical clustering plots
  lapply(reps, function(x) {
    pdf(file=paste0(out_folder, '/qc_', x, '_clustering.pdf'), h=8, w=6)
    plt = plotHC(dat_wide[[x]])
    draw(plt)
    dev.off()
  })

  getResultsFromLinearModel(dat_wide[['biorep_all']], folder)
})
```



# Old analysis

```{r}
z_scoring_results <- map(unique(iris$folder), function(folder) {
  out_folder = paste0('./output/', folder)
  z = iris %>% filter(folder == {{ folder }})

  # TODO: relying on the `numb` column here is not the best idea
  z1 = subset(z, grepl("10010201", numb)) %>% rename(AraIPTG01 = opacity)
  z2 = subset(z, grepl("1001001", numb)) %>% select(IPTG01 = opacity)

  z3 = cbind(z1, z2) %>%
    # Remove strains where opacity is less than min_colony_size_opacity in the IPTG plates (too
    # small colonies to draw any conclusions)
    filter(IPTG01 > min_colony_size_opacity) %>%
    # TODO: compare how this behaves when you also group by plate_replicate (to account for plate effects)
    # AND/OR compare this to within-plate gfp normalization beforehand (should have similar results)
    mutate(
      Z.IPTG    = scale(IPTG01),
      Z.AraIPTG = scale(AraIPTG01),
      # TODO: We should probably incoorporate a parameter to designate whether its TIC or TAC
      ## in z diff: for blockers is the difference between AraIPTG-IPTG, for triggers is IPTG-AraIPTG
      Z.diff    = Z.IPTG - Z.AraIPTG
    )

  z4 = z3 %>%
    group_by(genename) %>%
    summarize(
      Z.mean = mean(Z.diff, na.rm = T),
      Z.sd   = sd(Z.diff, na.rm = T),
      lower  = Z.mean - Z.sd,
      upper  = Z.mean + Z.sd
    ) %>%
    # clean the names
    mutate(
      gene.name = sub("[.].*", "", genename),
      gene.name = sub("_", ".", gene.name),
      gene.name = sub("gp", "", gene.name)
    ) %>%
    # drop the ones with SD is 'NA' i.e. genenames with a single replicate
    filter(!is.na(Z.sd)) %>%
    # call hits and add color helpers for plotting
    mutate(
      isHit = Z.mean > z_score_cutoff,
      ## for blockers we use blue "#377eb8ff", for triggers green
      color = ifelse(isHit, "green4", "grey65")
    )

  z_score_plot <-(dotplot = z4 %>%
    {
      ggplot(., aes(
        x = gene.name, y = Z.mean, ymin = lower, ymax = upper,
        fill = color, label = gene.name
      )) +
        geom_pointrange(pch = 21) +
        scale_fill_identity() +
        labs(
          title = folder,
          x = "Genes",
          y = "Triggering score"
        ) +
        geom_text_repel(data = filter(., isHit)) +
        theme_bw() +
        theme(
          text = element_text(size = 10),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.ticks.length = unit(.2, "cm")
        ) +
        geom_hline(yintercept = z_score_cutoff, linetype = "longdash", colour = "grey", size = 0.5)
    })

  ggsave(
    plot = z_score_plot,
    filename = str_c(out_folder, "/z_score_plot.pdf"),
    width = 9,
    height = 5
  )
  return(z4)
})
```
